---
title: A Simulation Exercise with the Exponential Distribution
author: Tim Wise
date: February 21, 2015
output: 
  html_document:
    keep_md: true

---

### Overview

### Simulations

 

####  The exponential distribution

The exponential distribution can be simulated in R with rexp(n, lambda) where lambda is the rate parameter. For this analysis, we set lambda = 0.2. The shape of the exponential distribution for lambda = 0.2 is shown below:
```{r}
hist(rexp(n=1000, rate=0.2))
```

The mean and variance of the exponential distribution are both 1 / lambda:
```{r}
lambda <- 0.2
mu <- 1 / lambda
sigma <- 1 / lambda
c(mu, sigma)
```

#### Generate 1000 sets each with 40 exponentials

Now we generate 1000 sets of 40 random exponential numbers. We store those sets in a matrix, M, with 1000 rows, one for each set, and 40 columns, one for each member of the set:
```{r}
n <- 1000
m <- matrix(data = rexp(n = n * 40, 
                        rate = lambda),
            nrow = n,
            ncol = 40)
str(m)
```

### Sample Means vs Theoretical Mean

We have learned in this class that:  
- the average of a random sample is itself a random variable, and  
- the distribution of that random variable is centered around the population mean, mu.

Let's see if we can show that for the exponential distribution. 

First, let's create a new random variable, Z, whose values are the average of a set of 40 exponentials. With the values in M, we can compute 1000 values for Z by taking the average of each row in M:
```{r}
summarize.rows <- 1
Z <- apply (M, summarize.rows, mean)
str(Z)
```
Now, let's compare the sample mean of Z with mu, the theoretical mean of our exponential distribution:
```{r}
round(c(mu, mean(Z)), 2)
```
We can see that the two compare very closely.

Let's visualize Z, looking at the distribution of values: 
```{r}
hist(Z)
```

The distribution of Z, the sample means, is centered around `r mu`, which is the theoretical mean of our exponential distribution. And the distribution of Z looks Gaussian (normally distributed) and quite a bit different from the exponential distribution we started with.

### Sample Variance versus Theoretical Variance

Now, 
Var(Z) = (sigma ** 2) / n


The theoretical variance of Z is:

```{r}
var_z <- (sigma^2)/1000
sd_z <- sqrt(var_z)
c(var_z, sd_z)

```
The sample variance is:
```{r}
var(Z)
sd(Z)
mean(Z)
length(Z)
var(Z)/1000
stderr <- sqrt(var(Z)/length(Z))
stderr
sd(Z)/sqrt(length(Z))

sigma^2/40
sigma^2/40
```

