---
title: "An Exercise in the Central Limit Theorem"
author: "Tim Wise"
date: "May 2015"
output:
  html_document:
    keep_md: yes
  pdf_document: null
---

### Overview

This is an exercise in the Central Limit Theorem (CLT) where we investigate the distribution of the average of exponential numbers and compare:  

- The sample mean to the theoretical mean
- The sample variance to the theoretical variance, and
- The sample distribution to the normal distribution

This exercise was a project for the Coursea class _Introduction to Statistical Inference_ which is part the _Data Science Specialization_ from Johns Hopkins University.

#### Note to graders: 

Since this was an exercise and a demonstration, rather than a formal paper or executive report, I chose to write this report in a narrative style interweaving discussion, code, and results. I think it is a good style for this exercise and let me practice the skills learned in the Reproducible Research class. The report is less than 6 pages, as per the rubric, but the code and details are not separated into an appendix. Grade it how you wish. Thanks.

### Process

For our purposes, the important points of the CLT, as summarized from the class notes, are as follows: 

- Consider a population of random numbers from any distribution with mean $\mu$ and variance $\sigma^2$.
- The average of $n$ random draws from this population is itself a random variable. This variable is commonly symbolized as $\bar X_n$.
- The mean of that random variable is the population mean, that is $E[\bar X_n] = \mu$.
- The variance of that random variable is $Var(\bar X_n) = \sigma^2/n$.
- The distribution of $\bar X_n$ is approximately standard normal, $N(\mu,\sigma^2/n)$.

In this exercise, we demonstrate those points of the CLT by doing the following: 

- We start with a population of random numbers from the exponential distribution.
- We create a new random variable that is the average of 40 random exponential numbers. In CLT terms, this random variable would be named  $\bar X_{40}$, but we will just call it $Z$.
- We generate 1000 values for $Z$. 
- We look at the distribution of $Z$ and compute its sample mean and variance.
- We compare the sample mean and variance to the theoretical values given to us by the CLT.
- We show that $Z$ is normally distributed.

### Simulations

First, let's create a random variable, $Z$, whose value is the average of 40 numbers drawn from an exponential distribution. We generate 1000 values for $Z$. It is the distribution of these values that we are interested in.

####  The exponential distribution

The exponential distribution is simulated in R with *rexp(n, lambda)* function where $n$ is the number of exponential numbers to generate and $lambda$ is the rate parameter (e.g., 35 web hits per minute). The mean and standard deviation of the exponential distribution are both $1/lambda$. 

For this analysis, we'll use $lambda = 0.2$. Let's set $lambda$ and compute the mean, $mu$, and the standard deviation, $sigma$:
```{r}
lambda <- 0.2
mu <- 1 / lambda
sigma <- 1 / lambda
c(lambda, mu, sigma)
```

Just for grins, let's look at the shape of an exponential distribution with a mean of 5:
```{r}
hist(rexp(n=1000, rate=1/5), main="Shape of an Exponential Distribution")
```

We see it's a highly skewed distribution, with a concentration of the values between 0 and the mean and a long tail extending above the mean. It looks most un-normal.

#### Random variable Z, the average of 40 exponential numbers 

Now, we generate the average of 40 random exponential numbers 1000 times. We store those values in a vector $Z$, which is our random variable of interest:
```{r}
# set the seed so others can reproduce our results
set.seed(123456789)
n <- 40
Z <- NULL
for (i in 1:1000) { 
  Z <- c(Z, mean(rexp(n, lambda))) 
}
str(Z)
```

### Sample Mean versus Theoretical Mean

The CLT states **the distribution of the average** of independent identically distributed (iid) variables **is normally distributed around the population mean**.

For our example, that implies the values of Z should be distributed around the mean of our exponential distribution, which is $mu=`r mu`$. Let's see if that's true. 

First, let's visualize $Z$, looking at the distribution of values: 
```{r}
hist(Z)
```

The distribution of $Z$, shown in the histogram above, certainly looks centered around $`r mu`$.

Now let's compute the sample mean of Z and compare with $mu$:
```{r}
theoretical_mean <- mu
sample_mean      <- mean(Z)

round(c(theoretical_mean, sample_mean), 2)
```
We can see that the two compare very closely, supporting the CLT claim.

### Sample Variance versus Theoretical Variance 

The CLT says the theoretical variance of $Z$ is $\sigma^2/n$, where $sigma$ is the standard deviation of our exponential distribution and $n$ is the number of exponential numbers we draw for each set.

Let's compare the theoretical variance of Z to the sample variance:

```{r}
theoretical_variance <- sigma^2/n
sample_variance      <- var(Z)

round(c(theoretical_variance, sample_variance), 2)
```
These two compare closely also, supporting the CLT claim.

### Distribution of Z compared to Standard Normal

Finally, the CLT states that, properly normalized, the distribution of $Z$ is standard normal. A **standard normal distribution** has a mean and median of 0 and a standard deviation of 1. Let's standardize Z and see if it compares favorably to the standard normal measures.

We normalize the values of $Z$ by subtracting $mu$, the theoretical mean of $Z$, and dividing by the theoretical standard deviation of $Z$ (the square root of the variance of $Z$):

$$\frac{Z - \mu}{\sigma / \sqrt{n}}$$

Let's normalize the $Z$ values and plot their distribution:
````{r}
normalizedZ <- (Z - mu)/(sigma / sqrt(n))
hist(normalizedZ)
````

The values certainly look normally distributed, centered around 0.

Now, let's compute the mean, median, and standard deviation of the normalized $Z$ values and compare them to the standard normal measures. Again, the standard normal distribution has these measures:
````{r}
standard_mean   <- 0
standard_median <- 0
standard_dev    <- 1

round(c(standard_mean, standard_median, standard_dev), 4)
````
Computing the measures for our normalized $Z$ values:
```{r}
normalizedZ_mean   <- mean(normalizedZ)
normalizedZ_median <- median(normalizedZ)
normalizedZ_dev    <- sd(normalizedZ)

round(c(normalizedZ_mean, normalizedZ_median, normalizedZ_dev), 4)
```
We see the mean and deviation of the normalized $Z$ are comparable (<2 %) to those of the standard normal distribution, however, the sample median is suspicious, off by almost 6%. So we conclude that $Z$ is _approximately_ normally distributed.

### Summary

In this exercise, we did a simulation to demonstrate the principles of the Central Limit Theorem. For a sample distribution of the average of 40 exponential numbers, we showed that:  

- The sample mean was comparable to the theoretical mean
- The sample variance was comparable to the theoretical variance, and
- The sample values are approximately normally distributed


